---
title: "Project Final Version"
author: "Carolina Payeras"
format: html
editor: visual
bibliography: references.bib
---

```{r, message=FALSE, echo=FALSE, warning=FALSE}
knitr::opts_chunk$set(message=FALSE, warning=FALSE, echo=FALSE)

```

# Introduction

The issue that this investigation tackles is something that many war-torn countries struggle with: first, how to end the conflict at hand, and secondly, how to reintegrate former combatants back into society? Once the violence has settled, this latter part becomes logistically and psychologically very difficult, as societies struggle to reckon with providing for both victims and perpetrators of conflict. These challenges associated with re-integration of a society post-war are the subject of numerous academic works, and this investigation aims to build off of existing discourse. In their comprehensive article “Civil War”, Blattman and Miguel explain that post-intrastate conflict, “victors and vanquished (and victims) are condemned to coexist in the same society, potentially exacerbating political and social divisions.” [@blattman_civil_2010] This could pose a challenge to any nation that prescribes a panacea of a solution to re-integration, as there could not only be emotional resistance towards ex-combatant cooperation with the state, as well as social fear and hostility towards the ex-fighters by civilians, especially those who were directly impacted by the violence. This investigation also builds off of work by Bara et.al, who propose investigating various forms of post-war violence discretely to gain a clearer picture of a society re-integrating their previous adversaries [@bara_civil_2021].

Furthermore, an article touching on this exact point of interest was recently authored by Julián Arévalo, who positions Colombia’s current negotiations within the wider context of how peace negotiations after violence should be studied, emphasizing that elements within negotiations, rather than just the act of negotiating itself, are pivotal to its success [@arevalo_complex_2025]. This is useful for having at back-of-mind with this investigation, as whether or not negotiations were shown to work could be whether or not the key steps of successful implementations of the negotiation were taken.

For over 50 years, Colombia's government has been engaged in a civil war with the guerrilla fighters known as the FARC (Revolutionary Armed Forces of Colombia), which has resulted in violence and fear around the country, as well as notorious drug-trafficking. In 2016, President Juan Manuel Santos signed a historic peace treaty with the FARC in the hopes of ending the conflict, with the FARC agreeing to disarm and transition into a legitimate political party and bringing about hope for a new Colombia. In 2018, however, successive President Iván Duque, who was a critic of the peace treaty, promised to 'modify' the treaty to be more punitive of former guerrillas, claiming the 2016 Treaty was too lenient. This change in position of the Colombian government only heightened tensions with the FARC, and the peace deal began to crumble as guerrilla forces resumed armed violence, and consequently the new branch of the FARC-EP gained momentum \[@gutierrez_d_toward_2020\].

The current Colombian government under President Gustavo Petro announced their policy of “Total Peace” in 2022, which proposed easing hostility towards the guerrillas in order to de-escalate violence and encourage negotiations and dialogue. However, after four years of the Petro administration, the positive effects of Total Peace and the enormous undertaking it has been for the government have been under scrutiny by organizations such as the Washington Office on Latin America, who claim that the size and scope of the project has overwhelmed the government and undermined the policy's success [@totalp2025].

Prior work has studied Colombia's peace negotiations in a more general light of whether they were effective in dis-empowering the violent potentially of armed groups such as the FARC, or now the powerful ELN (National Liberation Army), across the entire nation. Other scholars, such as Curiel et. al., studied the power of pro-democratic campaigns on re-building trust in democratic institutions among ex-FARC combatants, specifically targeting the tendency of the fighters to support ideologically hard-line attitudes, which had significant success in de-radicalizing former combatants [@curiel_democratic_2023]. If Petro's negotiation plan is unable to reduce violence in Colombia and encourage the guerrilla fighters to lay down weapons and trust in the state, Curiel's work would be a proven alternative route. However, given that violence seems to be continuing specifically within the Caribbean and rural areas of Colombia, this investigation offers to contribute more specifically to whether fewer people are losing their lives to guerrilla violence, regardless of the lost or gained power of these armed groups.

Although this investigation will not dive into the details of the plan itself, the language and manner in which Petro is approaching peace could also be integral to its success in decreasing the intensity of violence. Finally, measuring Total Peace’s success first via intensity of conflict rather than conflict gives us a chance to engage with smaller levels of success within a complex peace process rather than a crude view of violence trends in their entirety.

## Hypothesis

The question this paper aims to investigate is whether Gustavo Petro's Total Peace policy led to a decrease in curtailing civilian-targeted fatalities in states with histories of FARC activity. This paper hypothesizes that civilian-targeted fatalities will have decreased after the implementation of Total Peace. For this investigation, fatalities are used as a proxy for the intensity of violence rather than simply the occurrence of violence. I expect to see this relationship based on the policy being a large-scale government effort that aims to reduce hostility from the government towards the guerrillas, and the mechanism through which the policy would reduce violence towards civilians would be fewer state-FARC confrontations and less of a need to enforce dominance through killings. If the FARC responded violently when they felt their voices in government were being silenced and demonized, then maybe this effort would ease their concerns of ostracization and political disenfranchisement. This is not to say that it will overall civilian-targeted violence, which has actively been contested by scholars and institutions such as the ACLED (2024), but specifically it could likely reduce violent events and fatalities.

# Research Design

The data used for this investigation was gathered by the ACLED and published on the United Nations Office for the Coordination of Humanitarian Affairs' (OCHA) Humanitarian Data Exchange website. Specifically, the data focuses on incidences of violence against civilians per municipality, measured in events and fatalities from these violent incidences, on a weekly frequency. The ACLED defines civilian as “violence against civilians events and explosions/remote violence events in which civilians were directly targeted.” 

For the purposes of this investigation, we will only look at the number of incidences compositely on a year to year basis so that we can analyze the effectiveness of the policy and the trends that follow on a larger scale where change could be more significant, rather than a monthly or weekly basis. We will also look at the data by state, also known as departamentos, rather than by individual municipality for analysis purposes.

This investigation will only focus on eight departamentos that were identified by INDEPAZ as "focus areas", meaning the eight regions of the country that tend to struggle with the most armed violence, typically in the rural and Pacific regions but spread out across the country as well [@noauthor_informe_2023]. These are the areas that Total Peace theoretically would have the most impact if it were to be successful, which makes them useful proxies for whether Total Peace worked overall to decrease violence in the nation.

The variables from the data set that are used are Admin1 (departamentos), fatalities, events, and year. Yearly fatalities were calculated by summing weekly fatalities per year, and categorized by the *departamento* in which they took place; the unit of analysis therefore is departamento-year. The number of events and fatalities in the data per week per municipality frequently, but not always, match up to approximately a 1:1 ratio, which then begs the question whether it is necessary to choose either events or fatalities to focus on, and why. The means that frequency and intensity of the violence is quite correlated. This investigation will focus on fatalities because of the intention of the plan to de-escalate FARC activity, not necessarily stop it altogether. In addition, several works such as Bara.et al. focus on violence and its intensity in the broader discussion of re-integration, which for the purposes of contributing to this debate, this investigation will also measure fatalities.

The mean and the range of the variables over time on aggregate for the eight departamentos we are focusing on are below.

```{r, echo= FALSE, message=FALSE}
library(rio)
library(dplyr)
library(tidyverse)
library(readr)
library(haven)
library(purrr)
library(ggplot2)
library(readxl)
library(modelsummary)
library(fixest)

civilian_violence <- read_xlsx("input/civilian_raw.xlsx")
coca_production <- read_xlsx("input/coca_production_raw.xlsx")


# Only taking the variables I need from the greater data set (don't need postal code or country name, etc.)
civilian_violence1 <- civilian_violence |>
  select(Admin1, Admin2, Month, Year, Events, Fatalities)


# Now going to re-shape the data to make the unit of analysis departamento-year
civilian_violence2 <- civilian_violence1 |>
  group_by(Admin1, Year) |>
  summarise(
    total_events = sum(Events, na.rm = TRUE),
    total_fatalities = sum(Fatalities, na.rm = TRUE),
    .groups = "drop"
  ) |>
  rename("Departamento" = Admin1) |>
  filter(Year != 2026)


#Given that there are 32 different departamentos, for the purpose of this investigation I am onyl going to focus on the 8 regions with the highest historical number of guerrilla-based violent incidences that were indicated by INDEPAZ
# Using the %in% function to filter down cases where Departamento equals one of these names
civilian_violence3 <- civilian_violence2 |>
  filter(Departamento %in% c("Antioquia", "Arauca", "Cauca", "Norte de Santander", "Valle de Cauca", "Narino", "Bolivar", "Putumayo"))

civilian_violence3 <- civilian_violence3 |>
  mutate(Year = as.numeric(Year))

# Selecting only the key variables that I want for measuring fatalities over time per departamento
violence_fatalities <- civilian_violence3 |>
  select(Departamento, Year, total_fatalities)


mean_fatalities_per_year <- violence_fatalities |>
  group_by(Year) |>
  summarise(
    mean_fatalities = mean(total_fatalities, na.rm = TRUE),
    min_fatalities = min(total_fatalities, na.rm = TRUE),
    max_fatalities = max(total_fatalities, na.rm = TRUE))

library(knitr)

kable(mean_fatalities_per_year,
      col.names = c("Year", "Mean", "Min", "Max"),
      digits = 1,
      caption = "Mean fatalities per year across all regions")



```

## Main regression model

The regression model for this analysis models the relationship between time post-Total Peace, or post-2022, and the number of civilian-targeted fatilities. Assuming a linear model, we would use the following formula:

$$Violence_{dt} = \beta_0 + \beta_1 Post2022_t + \alpha_d + \epsilon_{dt}$$

The variables in the model are as follows:

- Violence = Number of violent events or fatalities
- t = time after Total Peace implementation in 2022 (in years)
- d = departamento
- Beta (0) = intercept of the model
- Beta (1) = the average difference in fatalities per *departamento*-year after 2022 relative to before 2022, within the same *departamento*
- alpha (d) = this variable controls for stable differences across departamentos such as population density, poverty, and regional economic output
- Edt = this is the error term for the model

Within this model, we want to focus on the value of Beta (1); should it be negative, this would confirm the hypothesis that the Total Peace Plan did decrease civilian-targeted fatalities in the chosen regions, even controlling for stable and time-dependent variation among the areas.

## Empirical Extension Proposal

One way that we can further validate that Total Peace is in fact working to decrease civilian-targeted fatalities is by seeing if coca production increased since Petro took office. If indeed Total Peace reduced fatalities through de-escalation in FARC-controlled areas, the coca cultivation in those areas should rise as well given increased government enforcement absence. In theory, less government oversight and patrol of FARC-controlled areas would mean higher coca production and fewer encounters with the state. If for some reason we see civilian fatalities dropping because of a factor besides Total Peace, such as economic development, then the coca development should not necessarily rise. The data that this extension will use comes from the United Nations Office on Drugs and Crime.

# Findings

After plotting the time-fatality data, it is clear that Total Peace has had very mixed effects when it comes to reducing the number of civilian fatalities.

```{r, echo= FALSE, message=FALSE}

## Going to visualize the the patterns of the number of civilian-targeted fatalities per region per year on a timescale
# Want t use the geom_vline function to add a vertical line that will mark when Total Peace was implemented

#| label: Figure 1
#| fig-cap: "Civilian-target fatalities by departamento from 2018 to 2025. President Petro's Total Peace policy was put into law in August of 2022, therefore the vertical line helps mark the trends in fatalities before and after the plan.

fatalities_plot <- ggplot(violence_fatalities, aes(x = Year, y = total_fatalities, color = Departamento, group = Departamento)) +
  geom_line(linewidth = 1) +
  geom_vline(xintercept = 2022, linetype = "dashed", color = "black") +
  labs(
    title = "Figure 1. Fatalities over time by region",
    x = "Year",
    y = "Number of fatalities",
    caption = "Figure. 1 demonstrates mixed results of fatalities per region after the implementation of total peace."
  ) +
  theme_minimal() +
  theme(plot.caption = element_text(hjust = 0, size = 9, face = "italic"))


print(fatalities_plot)

```

Figure 1 demonstrates that at the inital implementation of Total Peace by 2023, most regions had seen a decrease in the number of fatalities that year, even as the restrictions from the COVID-19 pandemic relaxed and more people were able to leave isolation. However, there seems to be increases from 2024 in certain regions such as Bolivar, Antioquia, and Norte de Santander started to increase in fatalities. This implies that Total Peace potentially only had short-term effects and is struggling to maintain lower levels of fatalities past 2024, which could indicate something amiss with the terms of negotiation that guerrilla groups may not agree with, or that decreased government oversight does in fact ultimately lead to more violence.

### Table 1. Effect of Total Peace on Civilian-Targeted Fatalities

```{r, echo=FALSE, message=FALSE, warning=FALSE}
#This is the code for running the regression model that I specified above
regression <- violence_fatalities |>
  mutate(
    post_2022 = if_else(Year > 2022, 1, 0))


#Here I am using the feols from the fixest package to include fixed effects estimations for the ordinary least squares regression that I am running, and the cluster means that it is clustering the standard errors by departamento for the group of years it contains
model_1 <- feols(
  total_fatalities ~ post_2022 | Departamento,
  cluster = ~Departamento,
  data = regression
)

modelsummary(model_1)

# I am using this technique to make my output look a bit cleaner so that there are not unnecessary rows, and to highlight the p-value and the estimation from my model
modelsummary(
  model_1,
  stars = TRUE,
  statistic = "p.value",
  coef_rename = c("post_2022" = "Post-Total Peace (2023+)"),
  gof_map = c("nobs", "r.squared"),
  title = "Effect of Total Peace on Civilian-Targeted Fatalities",
  notes = "Standard errors clustered at the departamento level. All models include departamento fixed effects."
)


```

Our model demonstrates a decrease of 7.69 (rounded to 7.7) fatalities per year after Total Peace was implemented, however the large p-value of 0.542 indicates that the results are statistically insignificant. One limitation of this particular model that could be leading to this insignificance is the number of departamentos that this investigation is using, as well as the only couple years of data available after the implementation of Total Peace.

## Empirical Extension

To further investigate the potential effectiveness of Total Peace on reducing civilian fatalities, as stated previously, we will look at whether coca production has risen in the areas in which there are high FARC activity.

```{r, echo=FALSE, message=FALSE}

coca_production_long <- coca_production |>
  pivot_longer(
    cols = starts_with("20"),   # selects all year columns
    names_to = "year",
    values_to = "hectares") |>
  mutate(year = as.integer(year)) |>
  select(year, hectares) |>
  mutate(hectares_thousand = hectares/1000) |>
  select(year, hectares_thousand)

coca_production_growth <- plot(coca_production_long$year, coca_production_long$hectares_thousand,
     type = "l",
     main = "Figure 2. Number of Hectares of Coca Cultivation in Colombia",
     xlab = "Year",
     ylab = "Hectares (thousands)")

print(coca_production_growth)

coca_production_long_2 <- coca_production_long|>
  mutate(years_post = year - 2010) |>
  select(years_post, hectares_thousand)

hectare_growth <- lm(hectares_thousand ~ years_post, data = coca_production_long_2)

summary(hectare_growth)


```

As seen in Figure 2, coca production does indeed continue to increase when the Total Peace plan is put into place, which shows that the mechanism of reducing fatalities via backing off of FARC territory. Although I could not locate exact departamento-level coca production breakdowns by year to compare across regions, a UNODC report from 2024 states that the increase in coca production in 2022 and 2023 come from typically FARC-controlled territories such as Cauca, Nariño, and Putumayo, indicating that it is safe to assume that the increase in coca production also aligns with less government oversight of these particular regions and overall armed activity.[@noauthor_colombia_nodate]

## Discussion and Policy Implications

The results from the model show potential for Total Peace to have succeeded in decreasing civilian fatalities in historically controlled FARC territory, however the very large p-value associated with the estimate from the model means that the result was not statistically significant. This aligns with the original visualization of the data, where Total Peace . Holistically, these results also map onto the arguments that Petro’s Total Peace plan has had mixed and sometimes very small effects if any, and also points to the difficulties that scholars frequently discuss of re-integrating combatants after war. This investigation concludes that the original hypothesis was far too simplistic in capturing the complexities of the Total Peace process that the results imply. One limitation of this study is the small number of departamentos that I used to calculate my results: narrowing down to only those that INDEPAZ focused on, this did decrease my "n" and therefore probably led to less validation of my results.

With the likely limited long-lasting effect of Total Peace, the analysis demonstrates that almost half of the regions where the policy had short-term results did not benefit from long-term reductions in fatalities. The election of the next Colombian president is coming soon in the next few months, and the legacy of Total Peace will either be carried on by one of the President's supporters, or rejected in favor of fighting fire with fire by politicians on the right. Either way, despite some success of Total Peace in decreasing violence in the most dangerous regions of the country, Colombia's struggle with intra-state conflict appears to be far from over.
